Performance Models for Data Transfers: A Case Study with Molecular Chemistry Kernels

### 2019-03-28 Article/article.tex
description of performance plots

\usepackage[utf8]{inputenc}
%%\usepackage{amsmath}
%%\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{threeparttable}
\usepackage{graphicx}
\usepackage{color}
\usepackage{paralist}
\usepackage{framed}
\usepackage{caption}
\usepackage{todonotes}
%%\newtheorem{theorem}{Theorem}
%%\newtheorem{corollary}{Corollary}[theorem]
%%\newtheorem{lemma}[theorem]{Lemma}
%%
%%\newtheorem*{example}{Example}
%%\newtheorem{remark}{Remark}

@@

@@

	We consider a machine called Cascade, available at PNNL, for our experiments. We obtain traces by running  two molecular chemistry applications,  double precision version of Hartree-Fock (HF) and Coupled Cluster Single Double (CCSD) of NWChem package with 4 Uracil molecules input on 10 nodes of this machine. Each node is composed of 16 Intel Xeon E5-2670 cores, hence we use 160 processes for the execution. Each process executes 300-800 tasks.  NWChem takes advantages of a Partitioned Global Address Space Programming Model, Global Arrays (GA), to use shared-memory programming APIs  on distributed memory computers.  Our data transfer model is quite simple and  we consider that all data transfers between each process's local memory and GA memory take the same route. Modeling of different routes of data transfers for the same source-destination pair, bandwidth sharing for different source-destination pairs and network congestion is more challenging and part of our future work. Our model is simple yet it provides insight to the application developers (or runtime system) that  in what-order all data transfer  are issued for the same source-destination pair to maximise communication-computation overlap. Our model is easily adaptable to any source-destination where there is one fixed route between source and destination (such as between CPU and GPU, one copy engine to transfer data from CPU (resp. GPU) to GPU (resp. CPU) ).
	\todo[inline]{More about tiling}
	Both application mainly perform two types of computations, tensor transpose and tensor contractions.  HF expects to specify a tile size and we set it to 100, so that each core can be efficiently utilize.  CCSD automatically determines tile sizes at different program points based on the input (molecule?).  Hence, HF operates on almost homogeneous tiles while CCSD on heterogeneous tiles.

@@

	\todo[inline]{Skumar: HF performance plots}
	\begin{figure}[htb]
		\includegraphics[scale=0.25]{../HFPlots/ratio_to_optimal.pdf}
		\caption{HF Performance.}
		\label{fig:staticOrderDynamicCorrectionsExample}
	\end{figure}
	\begin{figure}[htb]
	\includegraphics[scale=0.25]{../CCSDPlots/best_ccsd.pdf}
	\caption{HF best Performance.}
	\label{fig:staticOrderDynamicCorrectionsExample}
\end{figure}
	\todo[inline]{Skumar: CCSD performance plots}
		\begin{figure}[htb]
		\includegraphics[scale=0.25]{../CCSDPlots/ratio_to_optimal_ccsd.pdf}
		\caption{CCSD Performance.}
		\label{fig:staticOrderDynamicCorrectionsExample}
	\end{figure}
	\begin{figure}[htb]
		\includegraphics[scale=0.25]{../CCSDPlots/best_ccsd.pdf}
		\caption{CCSD best Performance.}
		\label{fig:staticOrderDynamicCorrectionsExample}
	\end{figure}
	\todo[inline]{Skumar: Effect of Random memory with HF or CCSD}

	\begin{figure}[htb]
	\includegraphics[scale=0.25]{../HFPlots/ratio_to_optimal_random_memory.pdf}
	\caption{HF Performance with random memory requirement of a task.}
	\label{fig:staticOrderDynamicCorrectionsExample}
\end{figure}	
	\todo[inline]{SKumar: Evaluate different metrics for HF, Cholesky and CCSD applications}
	\todo[inline]{SKumar: decide metrics: ratio to optimal, percentage overlap, 100 - percentage non overlap}
	\begin{itemize}
		\item Machine description
		\item temporary data requirement
		\item Description of all kernels (Molecular Chemistry Kernels)
		\item Think about results and graphs (Most probably results in $R$ plots)
	\end{itemize}
\newpage
\noindent The following table exhibits favorable scenarios for different heuristics. This table allows programmers to use the appropriate strategies to maximize communication-computation overlap for their applications.

\newpage
	\section{Conclusion and Perspectives}
	\begin{itemize}
		\item Evaluated different heuristics
		\item proved NP-complete problem
		\item Plan to implement the best heuristic in a runtime system  for NWChem algorithms and evaluation on Summit supercomputer
	\end{itemize}

>>>>

\usepackage[utf8]{inputenc}
%%\usepackage{amsmath}
%%\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{threeparttable}
\usepackage{graphicx}
\usepackage{color}
\usepackage{paralist}
\usepackage{framed}
%%\usepackage{caption}
\usepackage{subcaption}
\usepackage{todonotes}
%%\newtheorem{theorem}{Theorem}
%%\newtheorem{corollary}{Corollary}[theorem]
%%\newtheorem{lemma}[theorem]{Lemma}
%%
%%\newtheorem*{example}{Example}
%%\newtheorem{remark}{Remark}

@@

\subsection{Favorable Situation for Heuristics}
\noindent The following table exhibits favorable scenarios for different heuristics. This table allows programmers to use the appropriate strategies to maximize communication-computation overlap for their applications.

@@

	We consider a machine called Cascade, available at PNNL, for our experiments. We obtain traces by running  two molecular chemistry applications,  double precision version of Hartree-Fock (HF) and Coupled Cluster Single Double (CCSD) of NWChem package with 4 Uracil molecules input on 10 nodes of this machine. Each node is composed of 16 Intel Xeon E5-2670 cores, hence we use 160 processes for the execution. Each process executes 300-800 tasks.  NWChem takes advantages of a Partitioned Global Address Space Programming Model, Global Arrays (GA), to use shared-memory programming APIs  on distributed memory computers.  Our data transfer model is quite simple and  we consider that all data transfers between each process's local memory and GA memory take the same route. Modeling of different routes of data transfers for the same source-destination pair, bandwidth sharing for different source-destination pairs and network congestion is more challenging and part of our future work. Our model is simple yet it provides insight to the application developers (or runtime system) that  in what-order all data transfer  are issued for the same source-destination pair to maximise communication-computation overlap. Our model is easily adaptable to any source-destination where there is one fixed route between source and destination (such as between CPU and GPU, one copy engine to transfer data from CPU (resp. GPU) to GPU (resp. CPU) ).
	\todo[inline]{More about tiling}
	Both applications mainly perform two types of computations, tensor transpose and tensor contractions.  HF expects to specify a tile size and we set it to 100, so that each core can be efficiently utilize.  CCSD automatically determines tile sizes at different program points based on the input (molecule?).  Hence, HF operates on almost homogeneous tiles while CCSD on heterogeneous tiles.

@@

	\todo[inline]{may convert 10 subplots to 8 in all performance diagrams}
	\todo[inline]{increase font size in all plots, may be free scale-y}
	\todo[inline]{SKumar: decide metrics: ratio to optimal, percentage overlap, 100 - percentage non overlap}

	IWe evaluate different data transfer heuristics for the several memory capacities. From the obtained traces, first we determine the minimum requirement of the memory capacity $m_c$ to execute all tasks. Then we observe the behavior of all heuristics with memory capacity  $m_c$ to $2m_c$, in the gap of $0.1m_c$. Our performance metric is $ratio\_to\_optimal$, which is defined as the following. Let the makespan of a heuristic $H$ is $M_H$ and the makespan of the $optimal  case$ with infinite memory is $M_O$, then  $ratio\_to\_optimal (H)$ is $\frac{M_H}{M_O}$ (lower value is better). This value would be  greater or equal to $1$. A value closer to $1$ indicates that suitability of a particular heuristic and maximum possible communication-computation overlap has been achieved for that heuristic.

	Figures ~\ref{fig:ratio_to_optimal_hf} and ~\ref{fig:ratio_to_optimal_ccsd} depict the distribution of the performance  of each heuristic for all considered memory capacities, where plots are categorized by memory capacities. For each memory capacity and each heuristic, the box on the plot displays the median, first and last quartile, and the whiskers indicate minimum and maximum values, with outliers are shown by black dots.
\subsection{HF Performance}	
	\begin{figure}[htb]
		\includegraphics[scale=0.3]{../ExperimentalResults/ratio_to_optimal_hf.pdf}
		\caption{Comparison of different heuristics for HF.}
		\label{fig:ratio_to_optimal_hf}
	\end{figure}
 As indicated above, HF tasks are highly homogeneous, this is also noticeable in Figure~\ref{fig:ratio_to_optimal_hf}. All heuristics depict similar behavior for minimum memory capacity $m_c$  and  increasing the memory capacity slightly does not change the performance of all heuristics. As memory capacity increased further, dynamic variants of heuristics starts performing better and for the moderate memory capacities (close to $2m_c$ ), staticOrderDynamicCorrection variants outperforms others. \textit{Gilmore-Gomory} heuristic does not achieve good performance. it can be understand from the fact that task sequence is obtained considering no extra memory is available, but here we apply the same sequence for a different scenario,  overall memory consumed at each point is limited to memory capacity.

	\begin{figure}[htb]
	\includegraphics[scale=0.25]{../ExperimentalResults/inverse_ratio_to_optimal_hf-best.pdf}
	\caption{Comparision of best variants of all categories for HF.}
	\label{fig:ratio_to_optimal_best_hf}
	\end{figure}

Figure~\ref{fig:ratio_to_optimal_best_hf} shows the performance comparison of best variants of all categories with the $order_of_submission$ strategy. Static strategies are expected to perform better when there is not any memory capacity restriction, hence this plot indicates that static strategies face capacity bottleneck and underperform. Dynamic strategies achieve better performance with limited memory capacity and staticOrderDynamicCorrection  strategies suit in moderate memory capacity scenarios.

\subsection{CCSD Performance}

	\begin{figure}[htb]
	\includegraphics[scale=0.3]{../ExperimentalResults/ratio_to_optimal_ccsd.pdf}
	\caption{Comparison of different heuristics for CCSD.}
	\label{fig:ratio_to_optimal_ccsd}
	\end{figure}

CCSD application operates on tasks of different sizes, hence different heuristics depict distinct behaviors even at minimum memory capacity $m_c$ . Heterogeneity favors to dynamic strategies, therefore both dynamic and static+dynamic correction based strategies perform good than  static based strategies. Similar to HF,  static+dynamic correction based strategies outperform others as memory capacity becomes moderate.

	\begin{figure}[htb]
	\includegraphics[scale=0.25]{../ExperimentalResults/inverse_ratio_to_optimal_ccsd-best.pdf}
	\caption{Comparision of best variants of all categories for CCSD.}
	\label{fig:ratio_to_optimal_best_ccsd}
	\end{figure}

Figure~\ref{fig:ratio_to_optimal_best_ccsd} shows that best variants of dynamic and static+dynamic strategies achieve similar performance at minimum memory capacity $m_c$. But as memory capacity increases hterogeneity allows static+dynamic strategies to take advantage of static knowledge to get maximum overlap and dynamic correction to select another task in case of memory capacity limitation. Static strategies also start performing better at the end, which indicates that this application has potential for significant communication-computation overlap and pure dynamic strategies are unable to take scheduling decision based on this information.

\subsection{HF and CCSD Characteristics}

\begin{figure}[htb]
	\centering
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=.95\linewidth]{../ExperimentalResults/application_properties_hf.pdf}
		\caption{HF workloads}
		\label{fig:hfProperties}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=.95\linewidth]{../ExperimentalResults/application_properties_ccsd.pdf}
		\caption{CCSD workloads}
		\label{fig:ccsdProperties}
	\end{subfigure}
	\caption{HF and CCSD tasks characteristics.}
	\label{fig:ApplicationProperties}
\end{figure}

To get more insights into the considered applications, we also look in to their workloads. Figure~\ref{fig:ApplicationProperties} exhibits that HF is a communication intensive application and maximum 20\%  can be expected in the best scenario. While in CCSD , communications and computations are almost evenly distributed and  40-75\% communication-computation overlap is possible.

\subsection{LP Formulation}
%%\todo[inline]{SKumar: LP evaluation and discussion}
%%
%%	\todo[inline]{Skumar: Effect of Random memory with HF or CCSD}
%%	
\subsection{Impact of heterogeneity on different heuristics}

	\begin{figure}[htb]
	\includegraphics[scale=0.3]{../ExperimentalResults/ccsd-random/ratio_to_optimal_ccsd_random.pdf}
	\caption{Comparison of different heuristics for CCSD with highly heterogeneous tasks.}
	\label{fig:ratio_to_optimal_ccsd_random}
	\end{figure}

To know the affect of heterogeneity, we set the memory requirement of each task in CCSD trace randomly between 100 to 900 and observe the behavior of different heuristics in  Figure~\ref{fig:ratio_to_optimal_ccsd_random}. This plot also exhibits that static+dynamic variants make good use of static information to maximize communication-computation overlap and dynamic mechanism to minimize the penalty caused by memory capacity limitation. Pure static strategies suffer from memory capacity imitation and pure dynamic strategies are unable to maximize communication-computation overlap.

\subsection{Tasks in Batches}
%%	\begin{figure}[htb]
%%	\includegraphics[scale=0.25]{../ExperimentalResults/ccsd-random/ratio_to_optimal_ccsd_random.pdf}
%%	\caption{HF Performance with random memory requirement of a task.}
%%	\label{fig:staticOrderDynamicCorrectionsExample}
%%\end{figure}

\section{Conclusion and Perspectives}
\begin{itemize}
	\item Evaluated different heuristics
	\item proved NP-complete problem
	\item Plan to implement the best heuristic in a runtime system  for NWChem algorithms and evaluation on Summit supercomputer
\end{itemize}

%%	\begin{itemize}
%%		\item Machine description
%%		\item temporary data requirement
%%		\item Description of all kernels (Molecular Chemistry Kernels)
%%		\item Think about results and graphs (Most probably results in $R$ plots)
%%	\end{itemize}
